Assessing primitives performance\\ on multi-stage execution

### 2018-05-09 paper.tex
various fixes here and there, fixed refs

 \subsection{Pros and cons}

 - Requires recompilation of VM or plugin
 - Implementation language familiar to client dev
 - Implementation language familiar to vm dev
 - readability: number of lines of code ? 
 - Number of impl to maintain 1,2,3 or more if multiple clients.
 - Performance: stack switch, C performacne, cross file inlining, (InterpreterProxy indirection)
 - Control on generated code, diff llvm gcc

@@

\begin{table}
\centering
\begin{tabular}{cccccccc}
  \hline
    & \rot{recompilation needed} & \rot{familiar to client dev} & \rot{familiar to vm dev}} & \rot{number of LoC} & \rot{Implementations to maintain} & \rot{Performance} & \rot{Control over generated code} \\
  \hline
  Baseline    &             & \checkmark  &            & 10  & 3+ &  +  &  \\
   \hline
  SmartSyntax &             &  \checkmark &            & 20  & 3+ & ++  & -\\
   \hline
  SlangPlugin &             &  \checkmark &            & 35  & 2  & ++  & - \\
   \hline
  PureSlang   & \checkmark  &             & \checkmark & 45  & 1  & ++  & -\\
   \hline
  Slang+JIT   & \checkmark  &             & \checkmark & 85  & 1  & ++++ & + \\
  \hline
\end{tabular}
\caption{Trade-offs between implementations: summarized}
\label{tab:cmp_implem}
\end{table}

One or another implementation could be preferred depending on the intended aim: the main trade-off will be found between the engineering time and the performance. The table \ref{cmp_implem} summarizes the main pros and cons related to each implementation: the engineering time may tend to increase when more low-level languages are used, such as Cog's RLT for instance.
Also, a VM recompilation affects the overall complexity of the development process, but the resulting code might be faster to run. Likewise, the use of plugins may bring modularity, but as you will see in Section \ref{sec:eval}, the performance, although better than baseline version, could still be improved.

@@

The SlangPlugin, PureSlang and Slang+JIT implementations of this string comparison primitive have been recently (re)implemented in the Cog VM. In this section, we will compare the performances of these primitives against the other existing implementations performances. We have measures their respective execution times for different string sizes.

@@

The GCC-Linux evaluation was performed on an Asus Zenbook with Ubuntu 16.04 LTS, a 2,2 Ghz processor Intel Core 5, 8 Gb 1600MHz DDR3 of RAM. The Linux VM was compiled using GCC version 5.4.0.
The LLVM-Mac evaluation was performed on a MacBook pro with Mac OS 10.11.6, a 2.9Ghz processor Intel Core 5, 8 Gb 1867MHz DDR3 of RAM. The Mac VM was compiled using Apple LLVM version 8.0.0 (clang-800.0.38). 
The evaluation was performed on 32 bits VMs.

@@

   \subcaption{Size 1000}
   \end{subfigure}
   \subcaption*{\textbf{LLVM-Mac results}}
    \end{subfigure}
   \label{fig:bench}
\end{figure*}

@@

However, one could notice a rather striking performance difference between the 2 compilers: when compiling with GCC, we notive in Figure \ref{gcc} that the performances between the PureSlang and Slang+JIT implementations are really close for long strings (100 characters and longer), with respectively 24,42 and 24,48 relative speed-up factors for a 1000-characters string. Yet they greatly differ when LLVM is used, with respectively 13,44 and 26,40 relative speed-up factors.

@@

\begin{figure*}[htp!]
		\centering
		\includegraphics[width=1\linewidth]{figures/asm_cmp.pdf}
		\caption{GCC and LLVM: different optimizations on the comparison loop}
		\label{fig:Asm}
\end{figure*}

@@

\begin{code}
genPrimitiveStringCompareWith
	| instr jump jumpAbove jumpIncorrectFormat1 jumpIncorrectFormat2 jumpIncorrectFormat3 jumpIncorrectFormat4 jumpMidFailure jumpSuccess minSizeReg string1CharOrByteSizeReg string2CharOrByteSizeReg string1Reg string2Reg |
	<var: #jumpIncorrectFormat1 type: #'AbstractInstruction *'>
	<var: #jumpIncorrectFormat2 type: #'AbstractInstruction *'>
	<var: #jumpIncorrectFormat3 type: #'AbstractInstruction *'>
	<var: #jumpIncorrectFormat4 type: #'AbstractInstruction *'>
	<var: #jumpAbove type: #'AbstractInstruction *'>
	<var: #jumpSuccess type: #'AbstractInstruction *'>
	<var: #jump type: #'AbstractInstruction *'>
	<var: #jumpMidFailure type: #'AbstractInstruction *'>
	"I redefine those name to ease program comprehension"
	string1Reg := ReceiverResultReg.
	string2Reg := Arg0Reg.
	string1CharOrByteSizeReg := Arg1Reg.
	string2CharOrByteSizeReg := ClassReg.
	minSizeReg := SendNumArgsReg.
	"Load arguments in reg"
	cogit genLoadArgAtDepth: 0 into: string2Reg.
	"checks if string1 is a byteobject and get its size in bytes"
	self genGetFormatOf: string1Reg into: TempReg.
	cogit CmpCq: objectMemory firstByteFormat R: TempReg.
	jumpIncorrectFormat1 := cogit JumpLess: 0.
	cogit CmpCq: objectMemory firstCompiledMethodFormat R: TempReg.
	jumpIncorrectFormat2 := cogit JumpAboveOrEqual: 0.	
	self genGetNumSlotsOf: string1Reg into: string1CharOrByteSizeReg.
	(cogit LogicalShiftLeftCq: objectMemory shiftForWord R: string1CharOrByteSizeReg). 
	cogit AndCq: objectMemory wordSize - 1 R: TempReg R: TempReg. 
	cogit SubR: TempReg R: string1CharOrByteSizeReg. 
	"checks if string2 is a byteobject and get its size in bytes"
	self genGetFormatOf: string2Reg into: TempReg.
	cogit CmpCq: objectMemory firstByteFormat R: TempReg.
	jumpIncorrectFormat3 := cogit JumpLess: 0.
	cogit CmpCq: objectMemory firstCompiledMethodFormat R: TempReg.
	jumpIncorrectFormat4 := cogit JumpAboveOrEqual: 0.	
	self genGetNumSlotsOf: string2Reg into: string2CharOrByteSizeReg.
	(cogit LogicalShiftLeftCq: objectMemory shiftForWord R: string2CharOrByteSizeReg).
	cogit AndCq: objectMemory wordSize - 1 R: TempReg R: TempReg.
	cogit SubR: TempReg R: string2CharOrByteSizeReg.
	"Type and number of arguments are correct"
	"Compute the min"	 
	cogit CmpR: string1CharOrByteSizeReg R: string2CharOrByteSizeReg.
	jumpAbove := cogit JumpBelow: 0. 
	cogit MoveR: string1CharOrByteSizeReg R: minSizeReg. 
	jump := cogit Jump: 0. 
	jumpAbove jmpTarget: (cogit MoveR: string2CharOrByteSizeReg R: minSizeReg). 
	jump jmpTarget: (cogit CmpCq: 0 R: minSizeReg). 
	jumpSuccess := cogit JumpZero: 0. "if one of the string is empty, no need to go through the comparing loop"
	"Compare the bytes"
	cogit MoveCq: objectMemory baseHeaderSize  R: TempReg.
	cogit AddCq: objectMemory baseHeaderSize R: minSizeReg.
	instr := cogit MoveXbr: TempReg R: string1Reg R: string1CharOrByteSizeReg.
	cogit MoveXbr: TempReg R: string2Reg R: string2CharOrByteSizeReg.
	cogit SubR: string2CharOrByteSizeReg R: string1CharOrByteSizeReg. 
	jumpMidFailure := cogit JumpNonZero: 0. "the 2 compared characters are different, exit the loop"
	cogit AddCq: 1 R: TempReg.
	cogit CmpR: TempReg R: minSizeReg. 
	cogit JumpNonZero: instr.
	"all bytes from 1 to minSize are equal"
	self genGetNumBytesOf: string1Reg into: string1CharOrByteSizeReg.
	self genGetNumBytesOf: string2Reg into: string2CharOrByteSizeReg.
	jumpSuccess jmpTarget: (cogit SubR: string2CharOrByteSizeReg R: string1CharOrByteSizeReg).
	jumpMidFailure  jmpTarget: (cogit MoveR: string1CharOrByteSizeReg R: ReceiverResultReg).	
	self genConvertIntegerToSmallIntegerInReg: ReceiverResultReg.
	cogit genPrimReturn.
	jumpIncorrectFormat4 
		jmpTarget: (jumpIncorrectFormat3 
			jmpTarget: (jumpIncorrectFormat2 
				jmpTarget: (jumpIncorrectFormat1 jmpTarget: cogit Label))).
	^ CompletePrimitive
\end{code}

>>>>

@@

 In this section we compare the five different implementations according to the following criteria:
 \begin{itemize} 
 	\item \emph{Recompilation needed:} Diff no, plugin, full
	\item \emph{Implementation language familiar to developers:} Merged the 2 latter, explain ST dev ST, VM dev ST-Slang-C-ASM, maybe result is + ou - ? or + or 0
	\item \emph{Readability:} explain why number of lines of code. 
	\item \emph{Number of implementations to maintain:} explain fall-back code which is in all languages, Slang/C, JIT => maybe we ignore the fall-back code ?
	\item \emph{Stack switch from generated native code:} 
	\item \emph{Optimization potential:} discuss inlining C comp cross file and InterpreterProxy indirection
	\item \emph{Control over the generated code:} diff llvm gcc while JIT always the same
\end{itemize}

\begin{table}
\centering
\begin{tabular}{lccccccc}
  \hline
    & \rot{recompilation needed} & \rot{familiar to client dev} & \rot{familiar to vm dev} & \rot{number of LoC} & \rot{Implementations to maintain} & \rot{Performance} & \rot{Control over generated code} \\
  \hline
  Baseline    &             & \checkmark  &            & 10  & 3+ &  +  &  \\
   \hline
  SmartSyntax &             &  \checkmark &            & 20  & 3+ & ++  & -\\
   \hline
  SlangPlugin &             &  \checkmark &            & 35  & 2  & ++  & - \\
   \hline
  PureSlang   & \checkmark  &             & \checkmark & 45  & 1  & ++  & -\\
   \hline
  Slang+JIT   & \checkmark  &             & \checkmark & 85  & 1  & ++++ & + \\
  \hline
\end{tabular}
\caption{Trade-offs between implementations: summarized}
\label{tab:cmp_implem}
\end{table}

One or another implementation could be preferred depending on the intended aim: the main trade-off will be found between the engineering time and the performance. Table \ref{tab:cmp_implem} summarizes the main pros and cons related to each implementation: the engineering time may tend to increase when more low-level languages are used, such as Cog's RTL for instance.
Also, a VM recompilation affects the overall complexity of the development process, but the resulting code might be faster to run. Likewise, the use of plugins may bring modularity, but as you will see in Section \ref{sec:eval}, the performance, although better than baseline version, could still be improved.

@@

The SlangPlugin, PureSlang and Slang+JIT implementations of this string comparison primitive have been recently implemented in the Cog VM. In this section, we will compare the performances of these primitives against the other existing implementations performances. We have measures their respective execution times for different string sizes.

@@

\paragraph{GCC-Linux.}The GCC-Linux evaluation was performed on an Asus Zenbook with Ubuntu 16.04 LTS, a 2,2 Ghz processor Intel Core 5, 8 Gb 1600MHz DDR3 of RAM. The Linux VM was compiled using GCC version 5.4.0.

\paragraph{LLVM-Mac.}The LLVM-Mac evaluation was performed on a MacBook pro with Mac OS 10.11.6, a 2.9Ghz processor Intel Core 5, 8 Gb 1867MHz DDR3 of RAM. The Mac VM was compiled using Apple LLVM version 8.0.0 (clang-800.0.38).

\paragraph{Processor.}The following evaluation was performed on 32 bits Intel VMs (x86).

@@

   \subcaption{Size 1000}
   \end{subfigure}
   \subcaption*{\textbf{LLVM-Mac results}\vspace{-0.5cm}}
    \end{subfigure}
   \caption{String comparison benchmarks for strings of different sizes on different architectures}
   \label{fig:bench}
\end{figure*}

@@

However, one could notice a rather striking performance difference between the 2 compilers: when compiling with GCC, we notive in Figure \ref{fig:bench} that the performances between the PureSlang and Slang+JIT implementations are really close for long strings (100 characters and longer), with respectively 24,42 and 24,48 relative speed-up factors for a 1000-characters string. Yet they greatly differ when LLVM is used, with respectively 13,44 and 26,40 relative speed-up factors.

@@

\begin{figure*}[htp!]
		\centering
		\includegraphics[width=1\linewidth]{figures/asm_cmp.pdf}
		\caption{GCC and LLVM: different optimizations on the comparison loop \cba{I want another column with comments, what does each line do. I need some different representation, no addresses, labels for jumps, something standard would be nice.}}
		\label{fig:Asm}
\end{figure*}

@@

\begin{code}
genPrimitiveStringCompareWith
	| instr jump jumpAbove jumpIncorrectFormat1 jumpIncorrectFormat2 jumpIncorrectFormat3 jumpIncorrectFormat4 jumpMidFailure jumpSuccess minSizeReg string1CharOrByteSizeReg string2CharOrByteSizeReg string1Reg string2Reg |
	<var: #jumpIncorrectFormat1 type: #'AbstractInstruction *'>
	<var: #jumpIncorrectFormat2 type: #'AbstractInstruction *'>
	<var: #jumpIncorrectFormat3 type: #'AbstractInstruction *'>
	<var: #jumpIncorrectFormat4 type: #'AbstractInstruction *'>
	<var: #jumpAbove type: #'AbstractInstruction *'>
	<var: #jumpSuccess type: #'AbstractInstruction *'>
	<var: #jump type: #'AbstractInstruction *'>
	<var: #jumpMidFailure type: #'AbstractInstruction *'>
	"I redefine those name to ease program comprehension"
	string1Reg := ReceiverResultReg.
	string2Reg := Arg0Reg.
	string1CharOrByteSizeReg := Arg1Reg.
	string2CharOrByteSizeReg := ClassReg.
	minSizeReg := SendNumArgsReg.
	"Load arguments in reg"
	cogit genLoadArgAtDepth: 0 into: string2Reg.
	"checks if string1 is a byteobject and get its size in bytes"
	self genGetFormatOf: string1Reg into: TempReg.
	cogit CmpCq: objectMemory firstByteFormat R: TempReg.
	jumpIncorrectFormat1 := cogit JumpLess: 0.
	cogit CmpCq: objectMemory firstCompiledMethodFormat R: TempReg.
	jumpIncorrectFormat2 := cogit JumpAboveOrEqual: 0.	
	self genGetNumSlotsOf: string1Reg into: string1CharOrByteSizeReg.
	(cogit LogicalShiftLeftCq: objectMemory shiftForWord R: string1CharOrByteSizeReg). 
	cogit AndCq: objectMemory wordSize - 1 R: TempReg R: TempReg. 
	cogit SubR: TempReg R: string1CharOrByteSizeReg. 
	"checks if string2 is a byteobject and get its size in bytes"
	self genGetFormatOf: string2Reg into: TempReg.
	cogit CmpCq: objectMemory firstByteFormat R: TempReg.
	jumpIncorrectFormat3 := cogit JumpLess: 0.
	cogit CmpCq: objectMemory firstCompiledMethodFormat R: TempReg.
	jumpIncorrectFormat4 := cogit JumpAboveOrEqual: 0.	
	self genGetNumSlotsOf: string2Reg into: string2CharOrByteSizeReg.
	(cogit LogicalShiftLeftCq: objectMemory shiftForWord R: string2CharOrByteSizeReg).
	cogit AndCq: objectMemory wordSize - 1 R: TempReg R: TempReg.
	cogit SubR: TempReg R: string2CharOrByteSizeReg.
	"Type and number of arguments are correct"
	"Compute the min"	 
	cogit CmpR: string1CharOrByteSizeReg R: string2CharOrByteSizeReg.
	jumpAbove := cogit JumpBelow: 0. 
	cogit MoveR: string1CharOrByteSizeReg R: minSizeReg. 
	jump := cogit Jump: 0. 
	jumpAbove jmpTarget: (cogit MoveR: string2CharOrByteSizeReg R: minSizeReg). 
	jump jmpTarget: (cogit CmpCq: 0 R: minSizeReg). 
	jumpSuccess := cogit JumpZero: 0. "if one of the string is empty, no need to go through the comparing loop"
	"Compare the bytes"
	cogit MoveCq: objectMemory baseHeaderSize  R: TempReg.
	cogit AddCq: objectMemory baseHeaderSize R: minSizeReg.
	instr := cogit MoveXbr: TempReg R: string1Reg R: string1CharOrByteSizeReg.
	cogit MoveXbr: TempReg R: string2Reg R: string2CharOrByteSizeReg.
	cogit SubR: string2CharOrByteSizeReg R: string1CharOrByteSizeReg. 
	jumpMidFailure := cogit JumpNonZero: 0. "the 2 compared characters are different, exit the loop"
	cogit AddCq: 1 R: TempReg.
	cogit CmpR: TempReg R: minSizeReg. 
	cogit JumpNonZero: instr.
	"all bytes from 1 to minSize are equal"
	self genGetNumBytesOf: string1Reg into: string1CharOrByteSizeReg.
	self genGetNumBytesOf: string2Reg into: string2CharOrByteSizeReg.
	jumpSuccess jmpTarget: (cogit SubR: string2CharOrByteSizeReg R: string1CharOrByteSizeReg).
	jumpMidFailure  jmpTarget: (cogit MoveR: string1CharOrByteSizeReg R: ReceiverResultReg).	
	self genConvertIntegerToSmallIntegerInReg: ReceiverResultReg.
	cogit genPrimReturn.
	jumpIncorrectFormat4 
		jmpTarget: (jumpIncorrectFormat3 
			jmpTarget: (jumpIncorrectFormat2 
				jmpTarget: (jumpIncorrectFormat1 jmpTarget: cogit Label))).
	^ CompletePrimitive}
\end{code}
